A Logical Solution of the Linearity Challenge

### 2025-07-20 paper/calculus.tex
client server syntax and semantics

\begin{table}
    \caption{Syntax of \Calculus.}
    \label{tab:syntax}
    \begin{math}
        \displaystyle
        \begin{array}[t]{rcll}
            P, Q & ::= & \Link\x\y \\
                & | & \Fail\x \\
                & | & \Close\x \\
                & | & \Wait\x.P \\
                & | & \Select\x{\InTag_i}\z.P \\
                & | & \Case\x\z{P}{Q} \\
                & | & \Fork\x\y\z{P}{Q} \\
                & | & \Join\x\y\z.P \\
                & | & \Cut\x{P}{Q} \\
        \end{array}
    \end{math}
\end{table}

The calculus of pure sessions that is formalized in our development is dubbed
\Calculus and is closely related to the multiplicative, additive fragment of
CP~\citep{Wadler14}, without shared channels. Its syntax makes use of an
infinite set of \emph{channel names}, ranged over by $x$, $y$ and $z$, and is
shown in \cref{tab:syntax}.
%
A link $\Link\x\y$ denotes the merging of the $x$ and $y$ channels, so that
every message sent on one of the channels is forwarded to the other. This form
is useful for modeling \emph{free outputs}, as we will see later.\Luca{Mettere
esempio!}
%
The $\Close\x$ and $\Wait\x.P$ processes respectively model the output and input
of a termination signal on the channel $x$. The former process has no
continuation, while the latter process continues as $P$ once the session has
been terminated.
%
The $\Fail\x$ process denotes a runtime error concerning the session $x$.
%
The $\Select\x{\InTag_i}\z.P$ and $\Case\x\z{Q_1}{Q_2}$ processes respectively
denote the output and input of a label $\InTag_i$ along with a fresh
continuation $z$ on the channel $x$. Depending on the label $\InTag_i$, the
receiver continues as $Q_i$. In practice it is useful to consider arbitrary
(domain-specific) sets of tags instead of the canonical two $\InTag_1$ and
$\InTag_2$. This generalization has no substantial impact on the formalization,
so we stick to the classical case of ``binary choices'' in line with the usual
presentations of linear logic.
%
The $\Fork\x\y\z{P}{Q}$ and $\Join\x\y\z.R$ processes respectively denote the
output and input of a pair of fresh channels $y$ and $z$ on the channel $x$.
These forms are often used in conjunction with links to model \emph{free
outputs}, but can also be used to model bifurcating sessions whereby a
sequential protocol continues as two parallel sub-protocols.

The notions of free and bound names are fairly standard. Just bear in mind that
all action prefixes bind continuation channels in (some) continuation processes.
In particular, $\Select\x{\Tag_i}\z.P$ binds $z$ in $P$, $\Case\x\z{P}{Q}$ binds
$z$ in both $P$ and $Q$, $\Fork\x\y\z{P}{Q}$ binds $y$ in $P$ but not in $Q$ and
$z$ in $Q$ but not in $P$, and $\Join\x\y\z.P$ binds both $y$ and $z$ in $P$.
%
We write $\fn{P}$ for the set of channel names occurring free in $P$ and we
identify processes up to renaming of bound names.

@@

\begin{table}
    \caption{Operational semantics of \Calculus.}
    \label{tab:semantics}
    \[
        \begin{array}{rrcll}
        \SLink &
        \Link\x\y & \pcong & \Link\y\x \\
        \SComm &
        \Cut\x{P}{Q} & \pcong & \Cut\x{Q}{P} \\
        \SFail &
        \Cut\x{\Fail\y}{P} & \pcong & \Fail\y & x \ne y \\
        \SWait &
        \Cut\x{\Wait\y.P}{Q} & \pcong & \Wait\y.\Cut\x{P}{Q} & x \ne y \\
        \SSelect &
        \Cut\x{\Select\y{\InTag_i}\z.P}{Q} & \pcong & \Select\y{\InTag_i}\z.\Cut\x{P}{Q} & x \ne y,z \\
        \SCase &
        \Cut\x{\Case\y\z{P}{Q}}{R} & \pcong & \Case\y\z{\Cut\x{P}{R}}{\Cut\x{Q}{R}} & x \ne y,z \\
        \SForkL &
        \Cut\x{\Fork\y\u\v{P}{Q}}{R} & \pcong & \Fork\y\u\v{\Cut\x{P}{R}}{Q} & x \in \fn{P}\setminus\set{y,u,v} \\
        \SForkR &
        \Cut\x{\Fork\y\u\v{P}{Q}}{R} & \pcong & \Fork\y\u\v{P}{\Cut\x{Q}{R}} & x \in \fn{Q}\setminus\set{y,u,v} \\
        \SJoin &
        \Cut\x{\Join\y\u\v.P}{Q} & \pcong & \Join\y\u\v.\Cut\x{P}{Q} & x \ne y,u,v \\
        \\
        \RLink &
        \Cut\x{\Link\x\y}{P} & \red & P\subst\y\x \\
        % \RFail &
        % \Cut\x{\Fail\y}{P} & \red & \Fail\y & x \ne y \\
        \RClose &
        \Cut\x{\Close\x}{\Wait\x.P} & \red & P \\
        \RSelect &
        \Cut\x{\Select\x{\Tag_i}\z.P}{\Case\x\z{Q_1}{Q_2}} & \red & \Cut\z{P}{Q_i} & i\in\set{1,2} \\
        \RFork &
        \Cut\x{\Fork\x\y\z{P}{Q}}{\Join\x\y\z.R} & \red & \Cut\y{P}{\Cut\z{Q}{R}} \\
        \RCut &
        \Cut\x{P}{R} & \red & \Cut\x{Q}{R} & P \red Q \\
        \RCong &
        P & \red & Q & P \pcong R \red Q \\
        \end{array}
    \]
\end{table}

The operational semantics of \Calculus is shown in \cref{tab:semantics} and is
given by two relations: a \emph{structural pre-congruence} relation $\pcong$ and
a \emph{reduction} relation $\red$. Structural pre-congruence relates processes
that we consider to be essentially indistinguishable. Reduction relates
processes in which a communication occurs. In some presentations of linear
logic~\cite{Doumane}, these relations are respectively dubbed \emph{external}
and \emph{internal} reductions. Let us describe the two relations more in
detail.

@@

@@

\begin{table}
    \caption{Operational semantics of \Calculus.}
    \label{tab:semantics}
    \begin{mathpar}
        \inferrule[\LinkRule]{~}{
            \wtp{
                \Link\x\y
            }{
                x : A, y : \dual{A}
            }
        }
        \and
        \inferrule[\FailRule]{~}{
            \wtp{
                \Fail\x
            }{
                \Context, x : \Top
            }
        }
        \and
        \inferrule[\CloseRule]{~}{
            \wtp{
                \Close\x
            }{
                x : \One
            }
        }
        \and
        \inferrule[\WaitRule]{
            \wtp{
                P
            }{
                \Context
            }
        }{
            \wtp{
                \Wait\x.P
            }{
                \Context, x : \Bot
            }
        }
        \and
        \inferrule[\SelectRule]{
            \wtp{
                P
            }{
                \Context, y : A_i
            }
        }{
            \wtp{
                \Select\x{\Tag_i}\y.P
            }{
                \Context, x : A_1 \Plus A_2
            }
        }
        \and
        \inferrule[\CaseRule]{
            \wtp{
                P
            }{
                \Context, y : A
            }
            \\
            \wtp{
                Q
            }{
                \Context, y : B
            }
        }{
            \wtp{
                \Case\x\y{P}{Q}
            }{
                \Context, x : A \With B
            }
        }
        \and
        \inferrule[\ForkRule]{
            \wtp{
                P
            }{
                \ContextC, y : A
            }
            \\
            \wtp{
                Q
            }{
                \ContextD, z : B
            }
        }{
            \wtp{
                \Join\x\y\z{P}{Q}
            }{
                \ContextC, \ContextD, x : A \Ten B
            }
        }
        \and
        \inferrule[\JoinRule]{
            \wtp{
                P
            }{
                \Context, y : A, z : B
            }
        }{
            \wtp{
                \Join\x\y\z.P
            }{
                \Context, x : A \Par B
            }
        }
        \and
        \inferrule[\CutRule]{
            \wtp{
                P
            }{
                \ContextC, x : A
            }
            \\
            \wtp{
                Q
            }{
                \ContextD, x : \dual{A}
            }
        }{
            \wtp{
                \Cut\x{P}{Q}
            }{
                \ContextC, \ContextD
            }
        }
    \end{mathpar}
\end{table}

>>>>

\begin{table}
    \caption{Syntax of \Calculus.}\label{tab:syntax}
    \begin{tabular}{c}
        \begin{math}
            \displaystyle
            \begin{array}[t]{rcll}
                P, Q & ::= & \Link\x\y            & \text{link} \\
                    & | & \Fail\x                 & \text{fail} \\
                    & | & \Close\x                & \text{close} \\
                    & | & \Wait\x.P               & \text{wait} \\
                    & | & \Select\x{\InTag_i}\z.P & \text{select} \\
                    & | & \Case\x\z{P}{Q}         & \text{branch} \\
                    & | & \Fork\x\y\z{P}{Q}       & \text{fork} \\
                    & | & \Join\x\y\z.P           & \text{join} \\
                    & | & \Server\x\y.P           & \text{accept} \\
                    & | & \Client\x\y.P           & \text{request} \\
                    & | & \Weaken\x.P             & \text{weaken} \\
                    & | & \Contract\x\y\z.P       & \text{contract} \\
                    & | & \Cut\x{P}{Q}            & \text{cut} \\
            \end{array}
        \end{math}
    \end{tabular}
\end{table}

The calculus of pure sessions that is formalized in our development is dubbed
\Calculus and is a linear version of \CP~\citep{Wadler14,GayVasconcelos25}. Its
syntax makes use of an infinite set of \emph{channel names}, ranged over by $x$,
$y$ and $z$, and is shown in Table~\ref{tab:syntax}.
%
A link $\Link\x\y$ denotes the merging of the $x$ and $y$ channels, so that
every message sent on one of the channels is forwarded to the other. This form
is useful for modeling \emph{free outputs}, as we will see later.\Luca{Mettere
esempio!}
%
The $\Close\x$ and $\Wait\x.P$ processes respectively model the output and input
of a termination signal on the channel $x$. The former process has no
continuation, while the latter process continues as $P$ once the signa has been
received.
%
The $\Fail\x$ process denotes a runtime error concerning the channel $x$.
%
The $\Select\x{\InTag_i}\z.P$ and $\Case\x\z{Q_1}{Q_2}$ processes respectively
denote the output and input of a label $\InTag_i$ along with a fresh
continuation $z$ on the channel $x$. Depending on the label $\InTag_i$, the
receiver continues as $Q_i$. In practice it is useful to consider arbitrary
(domain-specific) sets of tags instead of the canonical two $\InTag_1$ and
$\InTag_2$. This generalization has no substantial impact on the formalization,
so we stick to the classical case of ``binary choices'' in line with the usual
presentations of linear logic.
%
The $\Fork\x\y\z{P}{Q}$ and $\Join\x\y\z.R$ processes respectively denote the
output and input of a pair of fresh channels $y$ and $z$ on the channel $x$. The
former process forks into $P$ and $Q$, each using $y$ and $z$ respectively. The
latter process continues as $R$. These forms are often used in conjunction with
links to model \emph{free outputs}, but can also be used to model bifurcating
sessions whereby a sequential protocol continues as two parallel sub-protocols.

Next we have process forms dealing with shared (non-linear) channels. The
$\Server\x\y.P$ and $\Client\x\y.P$ processes respectively denote \emph{servers}
and \emph{clients} acting on the $x$. Each request (from a client) spawns a copy
of the server's body using the continuation channel $y$. The process
$\Weaken\x.P$ denotes an explicit \emph{weakening}, that is a client that does
\emph{not} use the server waiting on $x$. The process $\Contract\x\y\z.P$
denotes an explicit \emph{contraction}, that is a client that uses the server
$x$ multiple times on the channels $y$ and $z$.

\emph{Cuts} of the form $\Cut\x{P}{Q}$ represent the parallel composition of $P$
and $Q$ connected by a channel $x$.

The notions of free and bound names are fairly standard. Just bear in mind that
all action prefixes bind continuation channels in (some) continuation processes.
In particular, $\Select\x{\Tag_i}\z.P$ binds $z$ in $P$, $\Case\x\z{P}{Q}$ binds
$z$ in both $P$ and $Q$, $\Fork\x\y\z{P}{Q}$ binds $y$ in $P$ but not in $Q$ and
$z$ in $Q$ but not in $P$, and $\Join\x\y\z.P$ binds both $y$ and $z$ in $P$.
%
Also, $\Server\x\y.P$ and $\Client\x\y.P$ bind $y$ in $P$ and
$\Contract\x\y\z.P$ binds $y$ and $z$ in $P$.
%
We write $\fn{P}$ for the set of channel names occurring free in $P$ and we
identify processes up to renaming of bound names.

@@

\begin{table}
    \caption{Operational semantics of \Calculus.}
    \label{tab:semantics}
    \begin{tabular}{@{}c@{}}
        \begin{math}
            \displaystyle
            \begin{array}{rrcll}
            \SLink &
            \Link\x\y & \pcong & \Link\y\x \\
            \SComm &
            \Cut\x{P}{Q} & \pcong & \Cut\x{Q}{P} \\
            \SFail &
            \Cut\x{\Fail\y}{P} & \pcong & \Fail\y & x \ne y \\
            \SWait &
            \Cut\x{\Wait\y.P}{Q} & \pcong & \Wait\y.\Cut\x{P}{Q} & x \ne y \\
            \SSelect &
            \Cut\x{\Select\y{\InTag_i}\z.P}{Q} & \pcong & \Select\y{\InTag_i}\z.\Cut\x{P}{Q} & x \ne y,z \\
            \SCase &
            \Cut\x{\Case\y\z{P}{Q}}{R} & \pcong & \Case\y\z{\Cut\x{P}{R}}{\Cut\x{Q}{R}} & x \ne y,z \\
            \SForkL &
            \Cut\x{\Fork\y\u\v{P}{Q}}{R} & \pcong & \Fork\y\u\v{\Cut\x{P}{R}}{Q} & x \in \fn{P}\setminus\set{y,u,v} \\
            \SForkR &
            \Cut\x{\Fork\y\u\v{P}{Q}}{R} & \pcong & \Fork\y\u\v{P}{\Cut\x{Q}{R}} & x \in \fn{Q}\setminus\set{y,u,v} \\
            \SJoin &
            \Cut\x{\Join\y\u\v.P}{Q} & \pcong & \Join\y\u\v.\Cut\x{P}{Q} & x \ne y,u,v \\
            \SClient &
            \Cut\x{\Client\y\z.P}{Q} & \pcong & \Client\y\z.\Cut\x{P}{Q} & x \ne y,z \\
            \SWeaken &
            \Cut\x{\Weaken\y.P}{Q} & \pcong & \Weaken\y.\Cut\x{P}{Q} & x \ne y \\
            \SContract &
            \Cut\x{\Contract\y\u\v.P}{Q} & \pcong & \Contract\y\u\v.\Cut\x{P}{Q} & x \ne y,u,v \\
            \SServer &
            \Cut\x{\Server\y\u.P}{\Server\x\v.Q} & \pcong & \Server\y\u.\Cut\x{P}{\Server\x\v.Q} & x \ne y,u \\
            \\
            \RLink &
            \Cut\x{\Link\x\y}{P} & \red & P\subst\y\x \\
            \RClose &
            \Cut\x{\Close\x}{\Wait\x.P} & \red & P \\
            \RSelect &
            \Cut\x{\Select\x{\Tag_i}\z.P}{\Case\x\z{Q_1}{Q_2}} & \red & \Cut\z{P}{Q_i} & i\in\set{1,2} \\
            \RFork &
            \Cut\x{\Fork\x\y\z{P}{Q}}{\Join\x\y\z.R} & \red & \Cut\y{P}{\Cut\z{Q}{R}} \\
            \RConnect &
            \Cut\x{\Server\x\y.P}{\Client\x\y.Q} & \red & \Cut\y{P}{Q} \\
            \RWeaken &
            \Cut\x{\Server\x\y.P}{\Weaken\x.Q} & \red & \Weaken{\seqof\z}.Q
            & \fn{P} = \set{y,\seqof\z} \\
            \RContract &
            \Cut\x{\Server\x\y.P}{\Contract\x{x'}{x''}.Q} & \red &
            \multicolumn{2}{l}{
                \Contract{\seqof\z}{\seqof\z'}{\seqof\z''}.\Cut{x'}{
                    \Server{x'}\y.P'
                }{
                    \Cut{x''}{
                        \Server{x''}\y.P''
                    }{
                        Q
                    }
                }
                ~^*
            }
            \\
            \RCut &
            \Cut\x{P}{R} & \red & \Cut\x{Q}{R} & P \red Q \\
            \RCong &
            P & \red & Q & P \pcong R \red Q \\
            \\
            \multicolumn{5}{c}{
                ~^*
                \fn{P} = \set{y,\seqof\z},
                P' = P\subst{\seqof\z'}{\seqof\z},
                P'' = P\subst{\seqof\z''}{\seqof\z}
            }
            \end{array}
        \end{math}
    \end{tabular}
\end{table}

The operational semantics of \Calculus is shown in Table~\ref{tab:semantics} and
is given by two relations: a \emph{structural pre-congruence} relation $\pcong$
and a \emph{reduction} relation $\red$. Structural pre-congruence relates
processes that we consider to be essentially indistinguishable. Reduction
relates processes in which a communication occurs. In some presentations of
linear logic~\cite{Doumane}, these relations are respectively dubbed
\emph{external} and \emph{internal} reductions. Let us describe the two
relations more in detail.

@@

The notion of \emph{duality} is standard and we write $\dual{A}$ for the dual of
$A$.

@@
